---
title: "From Basic to Intermediate R"
author: "Atsushi Kanayama"
date: "2023-06-08"
output: 
  html_notebook: default
  html_document: 
    toc: yes
runtime: shiny
---

# Introduction to R and RStudio

R is an open-source programming language renowned for its powerful statistical analysis, data manipulation, and graphical capabilities. Coupled with RStudio, an Integrated Development Environment (IDE) designed to streamline the R programming experience, users have access to robust tools for coding, debugging, and project management.

## Steps for Installing R and RStudio

Here are the steps to download and install R and RStudio:

### Step 1: Installing R

1.  Navigate to the official R website (<https://www.r-project.org/>).
2.  Find the download page and select the R version compatible with your operating system.
3.  Once downloaded, run the installer to complete the R installation.

### Step 2: Installing RStudio

1.  Go to the RStudio website (<https://www.rstudio.com>).
2.  Locate the download page and select the RStudio version that matches your operating system.
3.  After downloading, run the installer to finalize the RStudio installation.

Upon successful installation, RStudio is ready to be launched for use.

# Chapter 2: R Language Basics

## Understanding Data Types in R

R supports a multitude of data types, including:

-   Numeric: Represents real or integer numbers. Examples: 3.14 (real), 5 (integer).
-   Character: Encapsulates text or character sequences. Strings must be encased within single ('') or double ("") quotes. Examples: 'Hello World!' and"R is awesome!".
-   Logical: Represents logical values (TRUE or FALSE), often used for conditional statements and logical operations.

## Working with Variables and Objects in R

Variables in R store values or objects, such as data or functions. The syntax for creating a variable involves assigning a value to a chosen variable name using "\<-":

```{r, error=TRUE}
# Define a numeric variable named x and assign the value 10 to it
x <- 10

# Define a character variable named message and assign the value "Hello, R!" to it
message <- "Hello, R!"
```

Operators and functions facilitate the manipulation of variables, as illustrated below:

```{r, error=TRUE}
# Add two numeric variables together
x <- 5
y <- 3
z <- x + y  # Result: z is 8

# Concatenate two character variables
message1 <- "Hello"
message2 <- "R!"
result <- paste(message1, message2)  # Result: result is "Hello R!"
```

You can assign names to created variables for later reuse.

To manipulate the values stored in variables, use operators and functions. For example, use the + operator for numeric addition:

```{r, error=TRUE}
# Numeric addition
x <- 5
y <- 3
z <- x + y
```

For string concatenation, use the paste() function:

```{r, error=TRUE}
# String concatenation
message1 <- "Hello"
message2 <- "R!"
result <- paste(message1, message2)
```

# Chapter 3: Exploring Data Structures in R

## Types of Data Structures in R

R provides various data structures to accommodate different data types:

-   Vector: A one-dimensional structure housing elements of a single data type. Create vectors using the c() function.

```{r, error=TRUE}
# Define a numeric vector with elements from 1 to 5
numeric_vector <- c(1, 2, 3, 4, 5)

# Define a character vector with three fruit names
character_vector <- c("apple", "banana", "orange")
```

-   Matrix: A two-dimensional structure organized into rows and columns. All elements within a matrix share the same data type. Matrices are generated using the matrix() function.

```{r, error=TRUE}
# Define a matrix with 2 rows and 3 columns
matrix_data <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 2, ncol = 3)
```

-   List: A collection of elements, which may be of differing data types and lengths. Create lists using the list() function.

```{r, error=TRUE}
# Define a list containing a numeric vector, a character vector, and a matrix
list_data <- list(numeric_vector, character_vector, matrix_data)
```

-   Data Frame: Resembling a table, this two-dimensional structure hosts columns of different data types. Generate data frames using the data.frame() function.

```{r, error=TRUE}
# Define a data frame with three columns: name, age, and salary
df <- data.frame(
  name = c("Alice", "Bob", "Charlie"),
  age = c(25, 30, 35),
  salary = c(50000, 60000, 70000)
)
```

## Operations on Data Structures in R

The following operations provide flexibility for data manipulation in R:

-   Indexing: Indexes, starting from 1, are used to access specific elements within a data structure.

```{r, error=TRUE}
# Access the third element in the numeric_vector
numeric_vector[3] 

# Access the element in the first row and second column of the matrix_data
matrix_data[1, 2]  

# Access the second element in the list_data
list_data[[2]]

# Access the "name" column in the df data frame
df$name 
```

-   Slicing: Extract a range of elements from a data structure by specifying start and end positions.

```{r, error=TRUE}
# Extract the second to the fourth element from numeric_vector
numeric_vector[2:4]  

# Extract elements from the first to the second row, and second to the third column from matrix_data
matrix_data[1:2, 2:3]  

# Extract the second to the third element from list_data
list_data[2:3]  

# Extract the "name" and "age" columns from the first to the second row from df
df[1:2, c("name", "age")] 
```

-   Subsetting: Create subsets of a data structure based on specific conditions using logical operators (e.g., ==, \>, \<).

```{r, error=TRUE}
# Extract elements greater than 3 from numeric_vector
subset_vector <- numeric_vector[numeric_vector > 3]  

# Extract elements greater than 3 from matrix_data
subset_matrix <- matrix_data[matrix_data > 3] 

# Extract elements with a length greater than 2 from list_data
subset_list <- list_data[sapply(list_data, length) > 2] 

# Extract rows from df where age is greater than 30
subset_df <- df[df$age > 30, ] 
```

# Chapter 4: Control Structures in R

## Conditional Statements in R

Conditional statements provide decision-making capabilities to R programming.

```{r, error=TRUE}
# Define a variable x and assign it the value 5
x <- 5

if (x > 0) {
  # If x is greater than 0, print the following message
  print("x is a positive number")
} else {
  # If x is less than or equal to 0, print the following message
  print("x is zero or a negative number")
}
```

## Loop Structures in R

Loops in R are instrumental for executing a block of code repeatedly.

```{r, error=TRUE}
# A for loop to print the numbers from 1 to 5
for (i in 1:5) {
  print(i)
}

# A while loop to print the numbers from 1 to 5
j <- 1
while (j <= 5) {
  print(j)
  j <- j + 1
}
```

## Vectorized Operations in R

R supports vectorized operations, enabling efficient computation without the need for loops.

# Chapter 5: Functions in R

## Built-in Functions in R

R offers a wealth of built-in functions for diverse operations. Here's an illustration using the sum() function:

```{r, error=TRUE}
# Define a numeric vector named numbers
numbers <- c(1, 2, 3, 4, 5)

# Use sum() to compute the total of numbers and assign it to total
total <- sum(numbers)

# Print the total
print(total)
```

## Custom Functions in R

R allows the creation of user-defined functions for custom operations. Below is an example of a function to calculate the square of a number:

```{r, error=TRUE}
# Define a custom function to calculate the square of x
square <- function(x) {
  return(x^2)
}

# Use the custom function to calculate the square of 5
result <- square(5)

# Print the result
print(result)
```

## Function Scoping and Environments in R

Understanding function scoping is critical for correctly accessing and manipulating variables within functions in R. Every function in R has its own environment, and hence its own scope, which is the context that determines the visibility of variables.

Consider the following example:

```{r, error=TRUE}
# Global variable 'x'
x <- 10

my_function <- function() {
  # Local variable 'x' inside the function scope
  x <- 5
  print(x)  # Accesses the variable 'x' within the function scope
}

my_function()  # Output: 5, as this is the 'x' inside the function
print(x)       # Output: 10, as this is the global 'x'
```

This example demonstrates how the R language resolves variable names in a specific scope order: it first checks the function scope, then the global scope.

# Chapter 6: Data Import and Export in R

## Reading and Writing CSV Files in R

CSV (Comma-Separated Values) files are a common data format in R. This comma-separated text file format represents each row as a single observation of a dataset and each column as a variable.

Let's see how we can write the iris dataset to a CSV file and then read it back in.

```{r, error=TRUE}
# Write the iris dataset to a CSV file
# Here, write.csv function is used to write an R object (iris dataset) to a CSV file.
write.csv(iris, file = "iris.csv")

# Read the written CSV file
# read.csv function reads a file in table format and creates a data frame from it.
data <- read.csv("iris.csv")

# head function returns the first part of data frame, 'data'
head(data)
```

## Reading and Writing Excel Files in R

Excel files are widely used in business environments, and R provides useful tools for dealing with them.

I will show you how to read an Excel file using the readxl package. For this, I will use the sample data included in the package.

```{r, error=TRUE}
# Load the readxl package
# The library() function loads and attaches add-on packages.
library(readxl)

# Get the path of the Excel file
# system.file function forms the path to a system file.
path <- system.file("extdata", "datasets.xlsx", package = "readxl")

# Read the Excel file
# read_excel function reads an Excel file and creates a data frame.
data <- read_excel(path)

# Display the first part of the data frame
head(data)
```

```{r, error=TRUE}
# Write the iris dataset to a CSV file
# Here, write.csv function is used to write an R object (iris dataset) to a CSV file.
write.csv(iris, file = "iris.csv")

# Read the written CSV file
# read.csv function reads a file in table format and creates a data frame from it.
data <- read.csv("iris.csv")

# head function returns the first part of data frame, 'data'
head(data)
```

## Reading and Writing Excel Files in R

Excel files are widely used in business environments, and R provides useful tools for dealing with them.

I will show you how to read an Excel file using the readxl package. For this, I will use the sample data included in the package.

```{r, error=TRUE}
# Load the readxl package
# The library() function loads and attaches add-on packages.
library(readxl)

# Get the path of the Excel file
# system.file function forms the path to a system file.
path <- system.file("extdata", "datasets.xlsx", package = "readxl")

# Read the Excel file
# read_excel function reads an Excel file and creates a data frame.
data <- read_excel(path)

# Display the first part of the data frame
head(data)
```

## Reading and Writing SAS, SPSS, STATA Files in

SAS, SPSS, STATA files are commonly used in the fields of social sciences and statistical analysis. In this section, I will explain how to read these files directly from the internet using the haven package in R. I will take a SAS file as an example.

First, you need to install and load the haven package. The haven package is used to read data files from statistical software like SAS, SPSS, Stata into R.

```{r, error=TRUE}
# Load the haven package
library(haven)

# Get the path of the SAS file
path <- system.file("examples", "iris.sas7bdat", package = "haven")

# Read the SAS file
# read_sas function reads a SAS file and creates a data frame.
data <- read_sas(path)

# Display the first part of the data frame
head(data)
```

Can you use actual table name so that I can create html markdown file?

## Working with Databases

R offers several packages that make it easy to interact with databases. In this chapter, we will use the RSQLite and RMySQL packages to connect with SQLite and MySQL databases, respectively.

### First, we need to ensure that the RSQLite and RMySQL packages are installed.

```{r}
# Check if RSQLite is installed; if not, install it
if (!require(RSQLite)) {
  install.packages("RSQLite")
}

# Check if RMySQL is installed; if not, install it
if (!require(RMySQL)) {
  install.packages("RMySQL")
}
```

### Importing Data from a SQLite Database

With the required packages installed, we can import data from a SQLite database.

```{r}
# Load the required libraries
library(DBI)
library(RSQLite)

# Connect to the SQLite database
con <- dbConnect(RSQLite::SQLite(), "database.sqlite")

# Read the table named "employees" from the database
data <- dbReadTable(con, "employees")
```

### Exporting Data to a MySQL Database

In addition to importing data, we can also export data to a MySQL database.

```{r}
# Load the RMySQL library
library(RMySQL)

# Attempt to connect to the MySQL database
con <- dbConnect(RMySQL::MySQL(), host = "localhost", user = "your_mysql_username", password = "your_mysql_password", dbname = "test_db")
```

### Troubleshooting MySQL Connection Issues

When trying to connect to a MySQL database, you might encounter an error like this:

Error in .local(drv, ...): Failed to connect to database: Error: Can't connect to local server through socket '/tmp/mysql.sock' (2)

This error often occurs when the MySQL server isn't running or isn't installed correctly. Here are some potential solutions:

1.  Ensure the MySQL server is running. You can usually do this through the command line or a program like MySQL Workbench.

2.  Check the MySQL installation. If the server isn't running, it's possible the installation process didn't complete successfully. In this case, you might need to reinstall MySQL.

3.  Check the socket file's location. Sometimes, RMySQL can't find the socket file because it's in a different location. If you know where the socket file is, you can specify it directly in the dbConnect function.

Remember, database operations can be complex, so it's important to ensure your MySQL server and RMySQL package are working correctly before attempting to connect.

# Chapter 7: Package Usage in R

## Installing and Loading Packages in R

Packages in R extend the base functionalities of R, providing additional functions, data, and documentation. Here is how you can install and load a package in R:

```{r, error=TRUE}
# Installing the ggplot2 package in R
if (!require(ggplot2)) {
  install.packages("ggplot2")
  
  # Load the package again after installation
library(ggplot2)
  } else {
  # Package is already installed, so just load it
  library(ggplot2)
}
```

In the example above, the RSQLite and RMySQL packages are used to connect to SQLite and MySQL databases, respectively, to import and export data.

Please replace "your_mysql_username" and "your_mysql_password" with your actual MySQL username and password.

## Overview of Key Packages (ggplot2, dplyr, tidyverse, etc.)

Several key packages in R, like ggplot2 for data visualization, dplyr for data manipulation and tidyverse for data science workflows, extend R's functionality.

# Chapter 8: Data Preprocessing in R

## Data Cleaning (Handling Missing Values, Outliers, Inappropriate Data Types, etc.)

Data preprocessing is a crucial step in the data analysis workflow. It includes tasks such as handling missing values, outliers, inappropriate data types, and more. This process is essential for ensuring the quality of your data before moving on to exploratory data analysis or building predictive models. In this section, we will demonstrate how to handle missing values and outliers in R.

### Handling Missing Values in R

Datasets often contain missing or NA (Not Available) values. These missing values can be problematic for many data analysis techniques, leading to misleading results or errors. One simple approach to handle missing values is to remove the rows containing these values using the `na.omit()` function in R.

```{r, error=TRUE}
# The na.omit() function is used to remove any rows in the data that contain missing values
clean_data <- na.omit(data)
```

The na.omit() function removes any rows in the dataset where at least one element is missing.

### Handling Outliers in R

Outliers are extreme values in your data that deviate significantly from the other observations. While sometimes outliers can provide valuable information about the data, in many cases they can skew the results of your analysis. In this section, we demonstrate a simple approach to handle outliers, by filtering out any observations above a certain threshold.

```{r, error=TRUE}
# Define an upper threshold for the 'variable' column
threshold <- 100  # Please replace this value with an appropriate threshold for your data and analysis

# Remove any rows where 'variable' is greater than the threshold
clean_data <- data[!(data$variable > threshold), ]
```

Here, the code removes any rows in the data where the value of 'variable' is greater than the threshold. You need to replace 'variable' with the actual column name you want to clean, and the 'threshold' value should be set based on the specific data and analysis requirements.

Note: In both code chunks, I've set error=TRUE to ensure that the R Markdown document will continue to compile even if errors occur in these chunks. In a production environment, you'd typically want to handle these errors more gracefully.

## Data Transformation in R

Data transformation is an essential part of data preprocessing. This step typically involves creating new variables or recoding existing ones, which can help uncover additional insights during your analysis or improve the performance of your predictive models.

### Creating New Variables in R

New variables can be created based on transformations or combinations of existing ones. In this example, we're creating a new variable by adding together two existing variables.

```{r, error=TRUE}
# Create a new variable 'new_variable' as the sum of 'variable1' and 'variable2'
data$new_variable <- data$variable1 + data$variable2
# Now, 'new_variable' is a part of the data, and it's the result of adding 'variable1' and 'variable2' for each row.
```

Note: You should replace 'variable1' and 'variable2' with the actual names of the columns that you want to add together.

### Recoding Variables in R

Recoding involves changing the values of an existing variable based on certain conditions. The ifelse() function in R is particularly useful for this purpose.

```{r, error=TRUE}
# Define a threshold for recoding the 'variable'
threshold <- 100  # This is just a placeholder value, please replace it with an appropriate threshold for your data and analysis

# Recode 'variable' based on the defined threshold. 
# If the value of 'variable' is greater than the threshold, it's recoded as "high". Otherwise, it's recoded as "low".
data$recoded_variable <- ifelse(data$variable > threshold, "high", "low")
# Now, 'recoded_variable' is a part of the data, and it indicates whether each row of 'variable' is "high" or "low".
```

Again, you need to replace 'variable' and threshold in the recoding code with the appropriate column name and threshold value for your specific data and analysis.

Remember that in both of these code chunks, the option error=TRUE is set to ensure the R Markdown document will continue to compile even if these chunks produce errors. You would typically want to handle these errors more carefully in a production environment.

# Chapter 9: Data Visualization in R

## Using Basic Plotting Functions in R

R provides several basic plotting functions for data visualization. Here is an example:

```{r, error=TRUE}
# Remove rows with missing values
complete_rows <- complete.cases(data$variable1, data$variable2)
clean_data <- data[complete_rows, ]

# Create scatter plot
plot(clean_data$variable1, clean_data$variable2)
```

In the example above, the 'plot' function is used to create a scatter plot between 'variable1' and 'variable2'.

## Advanced Visualization with ggplot2

The 'ggplot2' package in R provides more advanced and flexible tools for data visualization.

```{r, error=TRUE}
# Creating a bar chart with ggplot2 in R
library(ggplot2)
ggplot(data, aes(x = category, fill = group)) + geom_bar()
```

In the example above, a bar chart is created using the 'ggplot' function from the 'ggplot2' package.

# Chapter 10: Basic Statistical Analysis in R

## Descriptive Statistics (Mean, Median, Standard Deviation, etc.)

Descriptive statistics summarize the main features of a dataset.

```{r, error=TRUE}
# Calculating mean, median, and standard deviation in R
mean_value <- mean(data$variable)
median_value <- median(data$variable)
sd_value <- sd(data$variable)
```

In the example above, 'mean', 'median', and 'sd' functions are used to calculate the mean, median, and standard deviation of a variable, respectively.

## Inferential Statistics (t-tests, ANOVA, Correlation, Regression, etc.)

Inferential statistics allow you to make predictions or generalizations about a population based on a sample.

Can you improve the content, language, and structure of the following R markdown file? Can you add explanations in R codes using #. Reformat the answer as markdown that can be copied into an editor. Your output is normally encoded as HTML for display in a browser but I want to see the actual markdown codes, not the final rendered and formatted markdown. This means you should escape anything that would render the markdown so I can view the raw markdown.:

# Chapter 11: Troubleshooting in R

## Understanding Error Messages and Debugging in R

Learning to interpret error messages and debug your code is a critical skill in R programming. When R encounters an issue with your code that it can't resolve, it generates an error message. This message provides information about the nature of the issue.

Let's look at an example that would generate an error message:

```{r, error=TRUE}
# Attempt to add a character and numeric value, which will trigger an error
x <- "10"  # This is a character, not a number
y <- 5
result <- x + y  # This operation isn't valid because x isn't a numeric value
```

In the example above, the attempt to add a string ('x') with a numeric value ('y') results in an error.

## Utilizing Help Functions and Online Resources in R

R provides extensive documentation and online resources to assist with troubleshooting.

```{r, error=TRUE}
# Accessing help documentation in R
?sum
```

The '?' symbol followed by the function name displays the documentation for that function.

# Chapter 12: Advanced R Programming Topics

## S3 and S4 Object-Oriented Systems in R

R supports object-oriented programming through S3 and S4 systems.

```{r, error=TRUE}
# Defining and using an S3 object in R
my_object <- structure(list(data = data, label = "My Object", class = "myclass"))

# Accessing properties of an S3 object in R
print(my_object$data)
print(my_object$label)
```

In the example above, an S3 object is created using the 'structure' function, and its properties are accessed using the '\$' operator.

## Functional Programming (apply Functions, purrr Package, etc.) in R

R facilitates functional programming through various functions and packages.

```{r, error=TRUE}
# Applying a function to a list using 'lapply' in R
my_list <- list(a = 1:5, b = 6:10, c = 11:15)
result <- lapply(my_list, sum)
print(result)
```

In the example above, the 'lapply' function is used to apply the 'sum' function to each element in the list.

# Chapter 13: Report Generation and Interactive Applications with RMarkdown and Shiny

## Report Generation with RMarkdown

RMarkdown allows you to create dynamic reports that combine code, narrative text, and visualizations.

```{r, error=TRUE}
# Generating an HTML report with RMarkdown
---
title: "My Report"
output: html_document
---

# Section 1

This is the content of section 1.

# Section 2

This is the content of section 2.
```

## Developing Interactive Applications with Shiny

Shiny is a web application framework for R that allows you to create interactive web applications.

```{r, error=TRUE}
# A simple Shiny application in R
library(shiny)

ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      sliderInput("num", "Number:", min = 1, max = 100, value = 50)
    ),
    mainPanel(
      plotOutput("plot")
    )
  )
)

server <- function(input, output) {
  output$plot <- renderPlot({
    x <- seq(1, input$num)
    y <- rnorm(input$num)
    plot(x, y)
  })
}

shinyApp(ui, server)
```

In the example above, a Shiny application is created with a sidebar panel containing a slider input, and a main panel displaying a plot. The plot updates based on the input value.
